### 2020-07-17 manuscript.tex
tidies

% Title must be 250 characters or less.
\begin{flushleft}
{\Large
\textbf\newline{Smarter dating moves and faster proposals: revisiting the phylogenetic relaxed clock model} % Please use "sentence case" for title and headings (capitalize only the first word in a title (or heading), the first word in a subtitle (or subheading), and any proper nouns).
}
\newline
% Insert author names, affiliations and corresponding author email (do not include titles, positions, or degrees).
\\
Jordan Douglas\textsuperscript{1,2*},
Rong Zhang\textsuperscript{1,2},
Alexei J. Drummond\textsuperscript{1,2,3},
Remco Bouckaert\textsuperscript{1,2}
\\
\bigskip
\textbf{1} Centre for Computational Evolution,  University of Auckland, Auckland, New Zealand\\
\textbf{2} School of Computer Science, University of Auckland, Auckland, New Zealand\\
\textbf{3} School of Biological Sciences, University of Auckland, Auckland, New Zealand
\\
\bigskip

@@

@@

Let $\mathcal{T}$ be a binary rooted time tree with $N$ taxa (and $2N-2$ branches). Let $L$ be the number of sites within the multiple sequence alignment $D$, and let $L_\text{eff}$ be the \textit{effective} number of sites in the alignment (ie. the number of site patterns). The posterior density of a phylogenetic model is described by

@@

for rate standard deviation $\sigma$ and other model parameters $\theta$. $\vec{\mathcal{R}}^{\,}$ is a vector of abstracted substitution rates, which is transformed into real rates by $r(\vec{\mathcal{R}}^{\,})$. Three methods of representing rates as $\vec{\mathcal{R}}^{\,}$ are presented in \textbf{\nameref{sect:rateparams}}.

Under the \textit{relaxed clock model}, each internal and leaf node is assigned a substitution rate $r_i = r(\mathcal{R}_i)$, which corresponds to its parent branch. There are a total of $|\vec{\mathcal{R}}^{\,}| = 2N-2$ rates, which are independently distributed under the relaxed clock model prior  \cite{drummond2006relaxed}.

%The substitution rate on branch $i$

\subsection*{Rate parameterisations}
\label{sect:rateparams}

@@

\subsubsection*{1. Real rates}
The natural (and unabstracted) parameterisation of a substitution rate is a real number $\mathcal{R}_i \in \mathbb{R}, \mathcal{R}_i > 0$ which is equal to the rate itself. Thus, under the \textit{real} parameterisation:

@@

Under the prior distribution $p(\vec{\mathcal{R}}^{\,} | \sigma)$, rates are often log-normally or exponentially distributed with a mean of 1:

\begin{align}
p(\mathcal{R}_i | \sigma) = \frac{1}{\mathcal{R}_i \sigma \sqrt{2\pi}} \exp \big( -\frac{(\ln \mathcal{R}_i - \mu)^2}{2\sigma^2} \big) \quad &(\text{ LogNormal}(\mu, \sigma)\;) \text{, or}  \\
p(\mathcal{R}_i | \sigma) = p(\mathcal{R}_i) = e^{-\mathcal{R}_i} \quad &(\text{ Exponential}(\lambda=1)\;)
\end{align}

where $\mu = -0.5\sigma^2$ is set such that the expected value of the log-normal distribution is 1.

Zhang and Drummond 2020 present a series of tree operators which propose internal/root node heights, and then recompute the rates of incident branches such that their genetic distances ($r_i \times \tau_i$) remain constant after the proposal. By maintaining genetic distances the likelihood can also be maintained. These operators account for the correlation which exists between branch rates and branch times -- a correlation which is induced by the likelihood function.

@@

The domain of $\vec{\mathcal{R}}^{\,}$ is uniformly distributed:

@@

This method was suggested in the original BEAST2 relaxed clock paper \cite{drummond2006relaxed} and has been widely used. However, the constant distance operators since introduced by Zhang and Drummond 2020 -- which are incompatible with the \textit{cat} parameterisation -- yield an increase in mixing rate under \textit{real} by up to an order of magnitude over that of \textit{cat}. Moreover, the implementation of the \textit{cat} model has come under criticism \cite{rannala2007inferring}.

@@

Transforming these quantiles into rates invokes the i-cdf of the underlying \textit{real} clock model distribution. Thus, while this approach has clear similarities with \textit{cat}, the domain of rates here is continuous (as opposed to being confined to a discrete number of bins) and is therefore compatible with the class of operators described by Zhang and Drummond 2020.

A potential disadvantage of the \textit{quant} method would be the computational requirements of continuously evaluating the i-cdf, especially for trees with large $N$. Hence, rather than evaluating the exact i-cdf $F^{-1}$, an approximation $\hat{F}^{-1}$ will be used instead:

@@

In this article we have extended \textit{quant} through a linear piecewise approximation of the i-cdf. As the piecewise approximation is linear, evaluating the derivatives $\frac{\partial}{\partial \mathcal{R}_i} \hat{F}^{-1}(\mathcal{R}_i) =  D \hat{F}^{-1}(\mathcal{R}_i)$ and $\frac{\partial}{\partial r_i} \hat{F}(r_i) = D \hat{F}(r_i)$  -- which are required for computing Hastings ratios -- is trivial. The approximation is comprised of $n$ pieces (where $n$ is fixed) and upper and lower rate boundaries $r_\text{min}$ and $r_\text{max}$. The approximation is displayed in \textbf{Fig \ref{fig:rateparams}} and further detailed in \textbf{\nameref{sect:S1Appendix}}.

Zhang and Drummond 2020 introduced several tree operators for the \textit{real} parameterisation -- including \textbf{Constant Distance}, \textbf{Simple Distance}, and \textbf{Small Pulley}. In this project, we extended these three operators so that they are compatible with the \textit{quant} parameterisation. These are presented in \textbf{\nameref{sect:S1Appendix}}.

@@

\begin{table}[h!]
\centering
\begin{tabular}{|l| l l l|} 
 \hline
 & Operator & Proposal & Parameters  \\
 1 & \textbf{Random walk operator} & $x^\prime \leftarrow x + s\Sigma$ & $\sigma, r, q$ \\
 2 & \textbf{Scale operator} & $x^\prime \leftarrow x \times e^{s\Sigma}$ & $\sigma, r$  \\
 3 & \textbf{Interval operator} & $\begin{array} {rl} y &\leftarrow \frac{u - x}{x - l} \times e^{s\Sigma} \\ x^\prime &\leftarrow \frac{u + l * y}{y + 1}  \end{array}$ & $q \;\; (l=0,u=1)$  \\
 4 & \textbf{Constant distance operators} & $x^\prime \leftarrow x + s\Sigma$ & $t$ \\

 \hline
\end{tabular}
\caption{Summary of proposal kernels $q(x^\prime|x)$ of clock model operators. In each operator, $\Sigma$ is drawn from either a $\text{Bactrian}(m)$ or $\text{Uniform}$ distribution (distributions are normalised so that they have a mean of 0 and a variance of 1). The scale size $s$ is tunable. The proposal kernel may apply to node heights $t$, clock standard deviation $\sigma$, clock rates $r$ (\textit{real} only), and clock rate quantiles $q$ (\textit{quant} only). The Scale operator acts on parameters with non-negative domains. The Interval operator proposes values which respect its domain ie. $l < x^\prime < u$. }
\label{table:kernels}
\end{table}

@@

The \textbf{Narrow Exchange} operator \cite{drummond2002estimating}, widely used in BEAST \cite{drummond2012bayesian,suchard2018bayesian} and BEAST2 \cite{bouckaert2019beast}, is similar to NNI, and works as follows (\textbf{Fig. \ref{fig:narrowexchange}}):

@@

\textul{\textit{Step 2}}. Identify the child of $E$ with the greater height. Denote this child as $D$ and its sibling as $C$ (ie. $t_D > t_C$).

@@

\begin{figure}[!h]
\includegraphics[width=\textwidth]{Figures/NarrowExchange.pdf}
\caption{\textbf{Depiction of Narrow Exchange and Narrow Exchange Rate operators.} Proposals are denoted by $\mathcal{T} \rightarrow \mathcal{T}^\prime$. The vertical axis corresponds to node height $t$. In the bottom figure, branch rates $r$ are indicated by line thickness. In this example, the $\mathcal{D}_{AE}$ and $\mathcal{D}_{CE}$ constraints are satisfied.}
\label{fig:narrowexchange}
\end{figure}

Lakner et al. 2008 \cite{lakner2008efficiency} found that tree operators which perterb topology (such as NNI and SPR) consistently perform better than those which also change branch lengths (such as LOCAL \cite{simon1998local} and Continuous Change \cite{jow2002bayesian}). If Narrow Exchange was adapted to the relaxed clock model by ensuring that genetic distances remain constant after the proposal, its performance may be improved even further. This may in turn permit proposing a new node height $t_D$ and therefore changing branch (time) lengths.

@@

Here, we present the \textbf{Narrow Exchange Rate} (NER) operator. Let $r_A$, $r_B$, $r_C$, and $r_D$ be the clock rates of nodes $A$, $B$, $C$, and $D$, respectively. In addition to the modest topological change applied by Narrow Exchange, NER also proposes new clock rates ${r_A}^\prime$, ${r_B}^\prime$, ${r_C}^\prime$, and ${r_D}^\prime$. While NER does not alter $t_D$ (ie. ${t_D}^\prime \leftarrow t_D$), we also consider NERw - a special case of the NER operator which embarks $t_D$ on a random walk:

@@

for random walk step size $s\Sigma$ where $s$ is a tunable scaler parameter and $\Sigma$ is drawn from a uniform or \textbf{\nameref{sect:randomwalks}}. NER (and NERw) are compatible with both the \textit{real} and \textit{quant} parameterisations. Analagous to the Constant Distance operator, new rates are proposed such that genetic distances between nodes $A$, $B$, $C$, and $E$ are maintained. Thus, there are $\binom{4}{2} = 6$ pairwise distance constraints.

@@

The total space of NER operators is comprised of all possible subsets of distance constraints (ie. $\{\},\{\mathcal{D}_{AB}\}, \{\mathcal{D}_{AC}\}, \dotso , \{\mathcal{D}_{AB}, \mathcal{D}_{AC}, \mathcal{D}_{AE}, \mathcal{D}_{BC}, \mathcal{D}_{BE}, \mathcal{D}_{CE} \}$) which are solvable. The simplest NER -- the null operator denoted by NER$\{\}$ -- does not satisfy any distance constraints. This is equivalent to Narrow Exchange.

@@

Computing the determinant $|J|$ invokes standard analytical differentiation and linear algebra libraries of MATLAB. 6 of the 54 operators were found to have $|J|=0$, corresponding to irreversible proposals, and were discarded.

\paragraph{3. Automated generation of BEAST2 operators.} Java class files are generated using string processing. Each class corresponds to a single operator, extends the class of a meta-NER-operator, and is comprised of the solutions found in \textbf{1} and the Jacobian determinant found in \textbf{2}. $|J|$ is further augmented if the \textit{quant} parameterisation is employed.

@@

A \textit{guided} operator incorporates knowledge about neighbouring states, while an \textit{adaptive} operator undergoes a training process to improve its efficiency over time \cite{roberts2007coupling}. In previous work, parsimony scores and conditional clade probabilities of neighbouring trees have been employed by guided tree operators \cite{hohna2012guided,zhang2019using,meyer2019adaptive} and the latter has also been explored as the basis of  adaptive tree operators \cite{hohna2012guided,meyer2019adaptive}. The (adaptive) mirror kernel \cite{thawornwattana2018designing} learns a target distribution which acts as a `mirror image' of the current point $x$.  The adaptable variance multivariate normal (AVMVN) kernel \cite{baele2017adaptive,suchard2018bayesian} learns correlations between parameters during MCMC. Baele et al. 2017  observed a large increase ($\approx 5-10 \times$) in sampling efficiency from using the AVMVN kernel on clock rates and substitution model parameters across partitions \cite{baele2017adaptive}.

@@

\begin{align}
	x^\prime &\leftarrow x + \sum\limits_{i=1}^N \sum\limits_{j=i}^N c_{i,j} \times s\Sigma \\
	\text {where }  c &= \texttt{cholesky} \{ (1-\beta) \frac{\Sigma_N}{N} + \beta \frac{\mathbb{I}_N}{N} \}.
\end{align}

The \texttt{cholesky}$\{Y\}$ decomposition returns a lower diagonal matrix $L$, with positive real diagonal entries, such that $Y = LL^\prime$ \cite{lindstrom1988newton, pourahmadi2007cholesky}. $s$ is a tunable step size parameter and $\Sigma$ is a random variable drawn from a proposal kernel (uniform or Bactrian for instance). Our BEAST2 implementation of the AVMVN kernel is adapted from that of BEAST \cite{suchard2018bayesian}.

@@

\textbf{2. Time to convergence}. Two independent MCMC chains are run and the time is measured until: a) the absolute difference in clade posterior probability between the two chains is less than 0.05 for all clades, b) the Rubin-Gelman statistic $\hat{R}$ \cite{gelman1992inference} of every estimated parameter is less than 1.05, and c) the effective sample size \cite{rambaut2018posterior} of every estimated parameter is greater than 200 in each chain. \\

\textbf{3. Mixing of parameters}. Key parameters are evaluated for the number of effective samples generated per hour (ESS/hr). \\

For the latter two criteria, methodologies are benchmarked using both simulated and empirical datasets -- the latter were compiled \cite{lanfear2019Github} and partitioned \cite{lanfear2016partitionfinder} by Lanfear et al. as `benchmark alignments' (\textbf{Table \ref{table:datasets}}). Each setting is run $10 \times$ per dataset and the average statistics across the $10 \times 2 = 20$ chains are reported. .

@@

Methodologies are benchmarked using the Intel(R) Xeon(R) Gold 6138 CPU (2.00 GHz), with two processing units per pair of MCMC chains.

@@

@@

We compared the three rate parameterisations described in \textbf{\nameref{sect:rateparams}}. All three settings use the standard BEAST2 clock model operators from Drummond et al. 2006 \cite{drummond2006relaxed}. \textit{real} and \textit{quant} additionally use the constant-distance tree operators described by Zhang and Drummond 2020. To determine whether the difference in performance between \textit{real}/\textit{quant} versus \textit{cat} is because of the constant-distance tree operators or the parameterisation itself, we also included benchmarked two additional settings: \textit{real 2006} and \textit{quant 2006}, which do not used the constant-distance operators. These five settings are validated in \textbf{\nameref{sect:WCSS_appendix}}.

%\textbf{Fig. \ref{fig:parameterisationResults}} shows that the \textit{real 2006} performs considerably worse than any of the other settings. This is due to the poor sampling of the prior under this setting (ie. low ESS of $p(\theta)$). The failure of \textit{real 2006} thus highlights the appeal of 1) the \textit{cat} or \textit{quant} parameterisations, both of which have trivial contributions to the prior density (ie. uniform priors), and 2) the smarter operators used by \textit{real} (Zhang and Drummond 2020). Due to its computational burden, \textit{real 2006} was not benchmarked for all of the datasets in \textbf{Table \ref{table:datasets}}.

%Our results show that the \textit{quant} parameterisation yields the best performance with respect to effective samples per hour. \textit{quant} outperforms \textit{quant 2006}, suggesting that the constant distance operators are effective. Furthermore, \textit{quant} outperforms \textit{real} especially at sampling from the posterior and prior distributions (ie. high ESS/hr for $p(\theta|D)$ and $p(\theta)$). This is most likely because of the uniform prior distribution of rate quantiles.

>>>>

% Title must be 250 characters or less.
\begin{flushleft}
{\Large
\textbf\newline{Smarter dating and faster proposals: revisiting the phylogenetic relaxed clock model} % Please use "sentence case" for title and headings (capitalize only the first word in a title (or heading), the first word in a subtitle (or subheading), and any proper nouns).
}
\newline
% Insert author names, affiliations and corresponding author email (do not include titles, positions, or degrees).
\\
Jordan Douglas\textsuperscript{1,2*},
Rong Zhang\textsuperscript{1,2},
Alexei J. Drummond\textsuperscript{1,2,3},
Remco Bouckaert\textsuperscript{1,2}
\\
\bigskip
\textbf{1} Centre for Computational Evolution,  University of Auckland, Auckland, New Zealand\\
\textbf{2} School of Computer Science, University of Auckland, Auckland, New Zealand\\
\textbf{3} School of Biological Sciences, University of Auckland, Auckland, New Zealand
\\
\bigskip

@@

The molecular clock hypothesis states that the evolutionary rates of biological sequences are approximately constant throughout time \cite{zuckerkandl1962molecular}.
This assumption forms the basis of phylogenetics, under which evolutionary trees and divergence dates of life forms are inferred from biological sequences, such as nucleotides and amino acids \cite{douzery2003local, drummond2006relaxed}.
In Bayesian phylogenetics, these trees and their associated parameters are estimated as probabilitiy distributions \cite{kuhner1995estimating, larget1999markov, mau1999bayesian}. 
Statistical inference is typically performed by sampling under the Metropolis-Hastings Markov chain Monte Carlo (MCMC) algorithm \cite{metropolis53, hastings70}.
Bayesian phylogenetic clock models have been implemented in many platforms including BEAST \cite{drummond2012bayesian}, BEAST2 \cite{bouckaert2019beast}, MrBayes \cite{ronquist2012mrbayes}, and RevBayes \cite{hohna2016revbayes}.

The simplest phylogenetic clock model -- the strick clock -- makes the convenient assumption that the evolutionary rate is constant across all lineages \cite{zuckerkandl1965evolutionary, kuhner1995estimating, larget1999markov}.  % Early examples? 
However, molecular substitution rates are known to vary over time, over population sizes, over evolutionary pressures, and over DNA replicative machineries \cite{gillespie1994causes, woolfit2009effective, loh2010optimization}.
Moreover, any given dataset may be clock-like (where the substitution rate has a small variance across lineages) or non clock-like (a large variance). 
In the latter case, a strict clock is probably not suitable.

This led to the development of relaxed (uncorrelated) clock models, under which each branch in the phylogenetic tree has its own substitution rate  \cite{drummond2006relaxed}.
Branch rates can be drawn from a range of probability distributions including log-normal, exponential, gamma, and inverse-gamma distributions \cite{drummond2006relaxed, lepage2007general, li2012model}.
This class of models is widely used, and has aided insight into many recent biological problems, including the 2016 Zika virus outbreak \cite{faria2017establishment} and the COVID-19 pandemic \cite{giovanetti2020first}.
% is compatible with increasingly prominent multispecies coalescent models \cite{ogilvie2017starbeast2},

Finally, although not the focus of this article, the class of correlated clock models assumes some form of auto-correlation between rates over time. 
The correlation itself can invoke a range of stochastic models, including compound Poisson \cite{huelsenbeck2000compound} and CIR processes \cite{lepage2007general}, or it can exist as a series of local relaxed clocks \cite{drummond2010bayesian}. 
However, due to the correlated and discrete nature of such models, the time until MCMC convergence may be cumbersome, particularly for larger datasets \cite{drummond2010bayesian}.

With the increasing availabilty of biological sequence data, the development of efficient Bayesian phylogenetic methods is more important than ever. 
The performance of MCMC is dependent not only on computational runtime but also the efficacy of an
MCMC setup to achieve its convergence.
A critical task therein lies the further advancement of MCMC operators.
Recent developments in this area include the advancement of guided tree proposals \cite{zhang2020using, meyer2019adaptive, hohna2012guided}, coupled MCMC \cite{altekar2004parallel, muller2019coupled},
adaptive multivariate transition kernels \cite{baele2017adaptive},
and other explorative proposal kernels \cite{yang2013searching, thawornwattana2018designing}.
In the case of clock models, smart tree proposals can account for correlations between substitution rates and divergence times \cite{zhang2020improving}. 
The rate parameterisation itself can also affect the ability to ``mix'' during MCMC \cite{drummond2006relaxed, li2012model, zhang2020improving}.

While a range of advanced operators and other MCMC optimisation methods have arisen over the years, there has yet to be a widescale performance benchmarking of such methods as applied to the relaxed clock model. 
In this article, we systematically evaluate how the relaxed clock model can benefit from i) different substitution rate parameterisations,  ii) the use of Bactrian proposal kernels \cite{yang2013searching}, iii) tree operators which account for correlations between substitution rates and times, and iv) adaptive multivariate operators \cite{baele2017adaptive}.
The discussed methods are implemented in and compared using BEAST2 \cite{bouckaert2019beast}.

% phylogenetic model -- and the operators used during MCMC -- to achieve convergence. 
%A great body of research of MCMC performance

% With increasing sequence availability, it is desirable to have faster MCMC
% This can be achieved through faster mixing and better performance (beagle)
% Address modern optimisation speedups - bactrian, AVMVN, + 2 others not covered in this paper including GUIDED tree proposals
% Address various attempts made to optimise clock models (Rong's paper, different parameterisations + 1 other not covered)
% While many ideas have come up over the years, a large scale systematic comparison of which methods yield faster mixing (if there is one) has not been applied

% In this article, benchmark a range of approaches including...

@@

Let $\mathcal{T}$ be a binary rooted time tree with $N$ taxa. Let $L$ be the number of sites within the multiple sequence alignment $D$, and let $L_\text{eff}$ be the effective number of sites in the alignment (i.e. the number of unique patterns across all sites in the alignment). The posterior density of a phylogenetic model is described by

@@

for substitution rate standard deviation $\sigma$ and other model parameters $\theta$. $\vec{\mathcal{R}}^{\,}$ is a vector of abstracted substitution rates, which is transformed into real rates by $r(\vec{\mathcal{R}}^{\,})$. Three methods of representing rates as $\vec{\mathcal{R}}^{\,}$ are presented in \textbf{\nameref{sect:rateparams}}.

Let $t_i$ be the height (time) of node $i$.
Every internal and leaf node $i$ in $\mathcal{T}$ is associated with a parental branch length (the height difference between $i$ and its parent) $\tau_i$ and a parental branch substitution rate $r_i = r(\mathcal{R}_i)$. 
There are a total of $2N-2$ internal and leaf nodes. 
Under the relaxed clock model, each $\mathcal{R}_i$ is independently distributed under the prior.

The posterior distribution is sampled by MCMC.
The probability of accepting proposed state $x^\prime$ from state $x$ is calculated using the Metropolis-Hastings ratio \cite{metropolis53, hastings70}:

\begin{eqnarray}
\label{eq:MCMC}
\alpha(x^\prime|x) =  \min\left( 1, \frac{p(x^\prime|D)}{p(x|D)} \frac{q(x|x^\prime)}{q(x^\prime|x)} \right).
\end{eqnarray}

$q(a|b)$ is the transition kernel: the probability of proposing state $b$ from state $a$.
The ratio between the two $\frac{q(x|x^\prime)}{q(x^\prime|x)}$ is known as the Hastings ratio \cite{hastings70}.

\subsection*{Substitution rate parameterisations}
\label{sect:rateparams}

@@

\clearpage
\subsubsection*{1. Real rates}
The natural (and unabstracted) parameterisation of a substitution rate is a real number $\mathcal{R}_i \in \mathbb{R}, \mathcal{R}_i > 0$ which is equal to the rate itself. Thus, under the \textit{real} parameterisation:

@@

Under the prior distribution $p(\vec{\mathcal{R}}^{\,} | \sigma)$, rates are log-normally distributed with a mean of 1:

\begin{align}
p(\mathcal{R}_i | \sigma) = \frac{1}{\mathcal{R}_i \sigma \sqrt{2\pi}} \exp \big( -\frac{(\ln \mathcal{R}_i - \mu)^2}{2\sigma^2} \big) 
\end{align}

where $\mu = -0.5\sigma^2$ is set such that the expected value of the log-normal distribution is 1.
In this article we only consider log-normal clock priors, however the methods discussed are general.

Zhang and Drummond 2020 present a series of tree operators which propose internal/root node heights, and then recompute the rates of incident branches such that their genetic distances ($r_i \times \tau_i$) remain constant after the proposal \cite{zhang2020improving}. 
These operators account for the correlation which exists between branch rates and branch times.
By keeping the genetic distance of each branch constant, the likelihood is unaltered by the proposal.

@@

These integers correspond to $n$ rate categories.
 The domain of $\vec{\mathcal{R}}^{\,}$ is uniformly distributed:

@@

This method has been widely used in BEAST and BEAST2 analyses \cite{drummond2006relaxed}. 
However, the recently developed constant distance operators  -- which are incompatible with the \textit{cat} parameterisation -- yield an increase in mixing rate under \textit{real} by up to an order of magnitude over that of \textit{cat} \cite{zhang2020improving}.

@@

Transforming these quantiles into rates invokes the i-cdf of the underlying \textit{real} clock model distribution. Thus, while this approach has clear similarities with \textit{cat}, the domain of rates here is continuous (as opposed to being confined to a discrete number of bins) and is therefore compatible with constant distance operators \cite{zhang2020improving}.

A potential disadvantage of the \textit{quant} method would be the computational requirements of continuously evaluating the i-cdf.
As the number of quantiles grows with the taxon count $N$, this drawback would be at its worst on large trees.
Hence, rather than evaluating the exact i-cdf $F^{-1}$, an approximation $\hat{F}^{-1}$ will be used instead:

@@

In this article we have extended \textit{quant} through a linear piecewise approximation of the i-cdf. As the piecewise approximation is linear, evaluating the derivatives $\frac{\partial}{\partial \mathcal{R}_i} \hat{F}^{-1}(\mathcal{R}_i) =  D \hat{F}^{-1}(\mathcal{R}_i)$ and $\frac{\partial}{\partial r_i} \hat{F}(r_i) = D \hat{F}(r_i)$  -- which are required for computing the Hastings ratios -- is trivial. The approximation is comprised of $n$ pieces (where $n$ is fixed) and upper and lower rate boundaries $r_\text{min}$ and $r_\text{max}$. The approximation is displayed in \textbf{Fig \ref{fig:rateparams}} and further detailed in \textbf{\nameref{sect:S1Appendix}}.

Zhang and Drummond 2020 introduced several tree operators for the \textit{real} parameterisation -- including \texttt{Constant Distance}, \texttt{Simple Distance}, and \texttt{Small Pulley}. In this project, we extended these three operators so that they are compatible with the \textit{quant} parameterisation. These are presented in \textbf{\nameref{sect:S1Appendix}}.

@@

\begin{table}[h!]
\centering
\begin{tabular}{|l| l l l|} 
 \hline
 & Operator & Proposal & Parameters  \\
 1 & \texttt{Random walk} operator & $x^\prime \leftarrow x + s\Sigma$ & $\sigma, r, q$ \\
 2 & \texttt{Scale} operator & $x^\prime \leftarrow x \times e^{s\Sigma}$ & $\sigma, r$  \\
 3 & \texttt{Interval} operator & $\begin{array} {rl} y &\leftarrow \frac{u - x}{x - l} \times e^{s\Sigma} \\ x^\prime &\leftarrow \frac{u + l * y}{y + 1}  \end{array}$ & $q \;\; (l=0,u=1)$  \\
 4 & \texttt{Constant distance} operator & $x^\prime \leftarrow x + s\Sigma$ & $t$ \\

 \hline
\end{tabular}
\caption{Summary of proposal kernels $q(x^\prime|x)$ of clock model operators.
\texttt{Random walk}, \texttt{Scale}, and \texttt{Interval} are standard BEAST2 operators.
 In each operator, $\Sigma$ is drawn from either a $\text{Bactrian}(m)$ or $\text{Uniform}$ distribution (distributions are normalised so that they have a mean of 0 and a variance of 1). The scale size $s$ is tunable. The proposal kernel may apply to node heights $t$, clock standard deviation $\sigma$, clock rates $r$ (\textit{real} only), and clock rate quantiles $q$ (\textit{quant} only). The Scale operator acts on parameters with non-negative domains. The Interval operator proposes values which respect the domain of the parameter i.e. $l < x, x^\prime < u$. }
\label{table:kernels}
\end{table}

@@

The \texttt{Narrow Exchange} operator \cite{drummond2002estimating}, widely used in BEAST \cite{drummond2012bayesian,suchard2018bayesian} and BEAST2 \cite{bouckaert2019beast}, is similar to NNI, and works as follows (\textbf{Fig. \ref{fig:narrowexchange}}):

@@

\textul{\textit{Step 2}}. Identify the child of $E$ with the greater height. Denote this child as $D$ and its sibling as $C$ (i.e. $t_D > t_C$).

@@

\begin{figure}[!h]
\includegraphics[width=\textwidth]{Figures/NarrowExchange.pdf}
\caption{\textbf{Depiction of \texttt{Narrow Exchange} and \texttt{Narrow Exchange Rate} operators.} Proposals are denoted by $\mathcal{T} \rightarrow \mathcal{T}^\prime$. The vertical axis corresponds to node height $t$. In the bottom figure, branch rates $r$ are indicated by line thickness. In this example, the $\mathcal{D}_{AE}$ and $\mathcal{D}_{CE}$ constraints are satisfied.}
\label{fig:narrowexchange}
\end{figure}

Lakner et al. 2008 \cite{lakner2008efficiency} found that tree operators which perterb topology (such as NNI and SPR) consistently perform better than those which also change branch lengths (such as LOCAL \cite{simon1998local} and Continuous Change \cite{jow2002bayesian}). If \texttt{Narrow Exchange} was adapted to the relaxed clock model by ensuring that genetic distances remain constant after the proposal, its performance may be improved even further. This may in turn permit proposing a new node height $t_D$ and therefore changing branch (time) lengths.

@@

Here, we present the \texttt{Narrow Exchange Rate} (NER) operator. Let $r_A$, $r_B$, $r_C$, and $r_D$ be the clock rates of nodes $A$, $B$, $C$, and $D$, respectively. In addition to the modest topological change applied by \texttt{Narrow Exchange}, NER also proposes new clock rates ${r_A}^\prime$, ${r_B}^\prime$, ${r_C}^\prime$, and ${r_D}^\prime$. While NER does not alter $t_D$ (i.e. ${t_D}^\prime \leftarrow t_D$), we also consider NERw - a special case of the NER operator which embarks $t_D$ on a random walk:

@@

for random walk step size $s\Sigma$ where $s$ is a tunable scaler parameter and $\Sigma$ is drawn from a uniform or \textbf{\nameref{sect:randomwalks}}. NER (and NERw) are compatible with both the \textit{real} and \textit{quant} parameterisations. Analagous to the \texttt{Constant Distance} operator, new rates are proposed such that genetic distances between nodes $A$, $B$, $C$, and $E$ are maintained. Thus, there are $\binom{4}{2} = 6$ pairwise distance constraints.

@@

The total space of NER operators is comprised of all possible subsets of distance constraints (i.e. $\{\},\{\mathcal{D}_{AB}\}, \{\mathcal{D}_{AC}\}, \dotso , \{\mathcal{D}_{AB}, \mathcal{D}_{AC}, \mathcal{D}_{AE}, \mathcal{D}_{BC}, \mathcal{D}_{BE}, \mathcal{D}_{CE} \}$) which are solvable. The simplest NER -- the null operator denoted by NER$\{\}$ -- does not satisfy any distance constraints. This is equivalent to \texttt{Narrow Exchange}.

@@

Computing the determinant $|J|$ invokes standard analytical differentiation and linear algebra libraries of MATLAB. 
6 of the 54 solvable operators were found to have $|J|=0$, corresponding to irreversible proposals, and were discarded.

\paragraph{3. Automated generation of BEAST2 operators.} Java class files are generated using string processing. Each class corresponds to a single operator, extends the class of a meta-NER-operator, and is comprised of the solutions found in \textbf{1} and the Jacobian determinant found in \textbf{2}. $|J|$ is further augmented if the \textit{quant} parameterisation is employed (\textbf{\nameref{sect:S1Appendix}}).

@@

A \textit{guided} operator incorporates knowledge about neighbouring states, while an \textit{adaptive} operator undergoes a training process to improve its efficiency over time \cite{roberts2007coupling}. In previous work, parsimony scores and conditional clade probabilities of neighbouring trees have been employed by guided tree operators \cite{hohna2012guided,zhang2020using,meyer2019adaptive} and the latter has also been explored as the basis of  adaptive tree operators \cite{hohna2012guided,meyer2019adaptive}. The (adaptive) mirror kernel \cite{thawornwattana2018designing} learns a target distribution which acts as a `mirror image' of the current point $x$.  The adaptable variance multivariate normal (AVMVN) kernel \cite{baele2017adaptive,suchard2018bayesian} learns correlations between parameters during MCMC. Baele et al. 2017  observed a large increase ($\approx 5-10 \times$) in sampling efficiency from using the AVMVN kernel on clock rates and substitution model parameters across partitions \cite{baele2017adaptive}.

@@

\begin{align}
	x^\prime &\leftarrow x + \sum\limits_{i=1}^N \sum\limits_{j=i}^N c_{i,j} \times s\Sigma \\
	\text {where }  c &= \text{cholesky} \left( (1-\beta) \frac{\Sigma_N}{N} + \beta \frac{\mathbb{I}_N}{N} \right).
\end{align}

The cholesky$(Y)$ decomposition returns a lower diagonal matrix $L$, with positive real diagonal entries, such that $Y = LL^\prime$ \cite{lindstrom1988newton, pourahmadi2007cholesky}. $s$ is a tunable step size parameter and $\Sigma$ is a random variable drawn from a proposal kernel (uniform or Bactrian for instance). Our BEAST2 implementation of the AVMVN kernel is adapted from that of BEAST \cite{suchard2018bayesian}.

@@

\textbf{2. Mixing of parameters}. Key parameters are evaluated for the number of effective samples generated per hour (ESS/hr). This calculation is performed by BEAST2 \cite{bouckaert2019beast}. \\

%\textbf{3. Time to convergence}. Two independent MCMC chains are run and the time is measured until: a) the absolute difference in clade posterior probability between the two chains is less than 0.05 for all clades, b) the Rubin-Gelman statistic $\hat{R}$ \cite{gelman1992inference} of every estimated parameter is less than 1.05, and c) the effective sample size \cite{rambaut2018posterior} of every estimated parameter is greater than 200 in each chain. \\

Methodologies are benchmarked using one simulated and nine empirical datasets -- the latter were compiled \cite{lanfear2019Github} and partitioned \cite{lanfear2016partitionfinder} by Lanfear et al. as `benchmark alignments' (\textbf{Table \ref{table:datasets}}).
To facilitate convergence on complex datasets, each posterior distribution is sampled using coupled MCMC (MC$^3$) with 4 chains \cite{muller2019coupled}.
Each methodology is benchmarked across 5 replicates on each dataset, using an the Intel Xeon Gold 6138 CPU (2.00 GHz).

@@

@@

\subsection*{Evaluating constant distance operators}

@@

We compared the three rate parameterisations described in \textbf{\nameref{sect:rateparams}}. All three settings use the standard BEAST2 clock model operators from Drummond et al. 2006 \cite{drummond2006relaxed}. \textit{real} and \textit{quant} additionally use the constant-distance tree operators \cite{zhang2020improving}. To determine whether the difference in performance between \textit{real}/\textit{quant} versus \textit{cat} is because of the constant-distance tree operators or the parameterisation itself, we also included benchmarked two additional settings: \textit{real 2006} and \textit{quant 2006}, which do not used the constant-distance operators. These five settings are validated in \textbf{\nameref{sect:WCSS_appendix}}.

%\textbf{Fig. \ref{fig:parameterisationResults}} shows that the \textit{real 2006} performs considerably worse than any of the other settings. This is due to the poor sampling of the prior under this setting (i.e. low ESS of $p(\theta)$). The failure of \textit{real 2006} thus highlights the appeal of 1) the \textit{cat} or \textit{quant} parameterisations, both of which have trivial contributions to the prior density (i.e. uniform priors), and 2) the smarter operators used by \textit{real} (Zhang and Drummond 2020). Due to its computational burden, \textit{real 2006} was not benchmarked for all of the datasets in \textbf{Table \ref{table:datasets}}.

%Our results show that the \textit{quant} parameterisation yields the best performance with respect to effective samples per hour. \textit{quant} outperforms \textit{quant 2006}, suggesting that the constant distance operators are effective. Furthermore, \textit{quant} outperforms \textit{real} especially at sampling from the posterior and prior distributions (i.e. high ESS/hr for $p(\theta|D)$ and $p(\theta)$). This is most likely because of the uniform prior distribution of rate quantiles.
